---
title: "Chapter 1: Data Visualization"
format: html
execute:
  warning: false
  message: false
  echo: false
---

#### Intro to Data Viz
The book begins with a basic description of data Science as a process. The core of it is what Wickham calls "an iterative cycle of **transforming**, **viziualizing**, and **modeling**." This is preceeded by **importing**, and then **tidying** data (*tidy format: columns are variables and rows are observations*). Once a point of understanding is reached in the core process, your findings are **communicated**. 

He also communicates that **tidying** and **transforming** have a relationship that, as a joint action, are called **wrangling**. 

Another concept is the difference between a **dataframe** and was in colloquially called a "dataset". Essentially, a **dataframe** is one spreadsheet. A dataset may be an excel sheet with tabs. Each of those tabs is a **dataframe**. 

<!-- 1.2.1 -->
```{r penguins}
library(palmerpenguins)
# palmerpenguins is a dataset stored in an R package
# It is also probably the most widely used dataset in R for examples, development, and teaching
# comments can exist in chunks. I don't like showing code in knitted files. 
```

```{r tidyverse}
library(tidyverse)
# tidyverse is most of the toolset for modern Data Science in R
```

<!-- 1.2.2 -->

```{r glimpse}
#| echo: true
glimpse(penguins)
```

The `penguins` dataframe comes from the `palmerpenguins` package. It is tidy out-of-the-box.

#### Types of Tables
###### There's levels to this

`ggplot()` is the TidyVerse command to start making a vizualization. If you enter the absolute basics of the ggplot() function you will get a blank slate. 

To make a scatterplot, you set up a ggplot and stick a geom_point() on top it. 

```{r}
#| label: fig-scatter-basic
#| fig-cap: ""
ggplot( # Written this way, I interpret the basics of a ggplot as settings in the `ggplot()` command. 
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g))  +  # at this point, there is a "live" ggplot seed
  geom_point() # And then `geom_point()` interprets the ggplot seed in it's own way. 
```


<!-- This will spit out an error in the console:

Removed 2 rows containing missing values or values outside the scale
range (`geom_point()`).  

That is because the penguins dataset has 2 missing values-->

This basic scatterplot works, but it looks terrible. The fun of R is the granularity you have to create exactly what you want in a vizualization. 

<!--  -->

The book dips the readers' toe into things by showing them labels, color and shape customization, and how to vizualize a 2 factor linear model, in the form of a trend line (with degrees of freedom to boot.)


```{r ggthemes}
library(ggthemes)
```
```{r}
#| label: fig-scatter-book
#| fig-cap: ""
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

But I have a lot of problems with this admittedly improved plot. mapping species to both color and shape is redundant. 

```{r}
#| label: fig-scatter-refined
#| fig-cap: ""
penguins |> # my preferred way of writing ggplots is to start with the dataframe and pipe it into ggplot. I think of it as "plot of penguins" like this, instead of "plot made of penguins".
    ggplot(mapping = aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(mapping = aes(color = species),
                             size = 3) + # set size of point manually
    geom_smooth(method = "lm",
                color = "black", 
                se = FALSE, # Turn of standard error
                fullrange = TRUE) + # aesthetic choice to make the line extand to the edge. 
    scale_color_viridis_d(option = "magma", begin = 0.2, end = 0.8) + # my preferred viridis settings that I came to throught trial and error in the past. 
    labs(
        x = "Flipper Length (mm)",
        y = "Body Mass (g)",
        title = "Gentoos are huge", # description of a finding
        caption = "Source: palmerpenguins" # always site source
    ) +
    scale_x_continuous(limits = c(171, 233), expand = expansion(0)) +
    scale_y_continuous(labels = scales::label_number(),
                       expand = expansion(mult = 0.02)) +
    theme_few() +
    theme(
      axis.title = element_text(size = rel(1.25)),
      axis.text = element_text(size = rel(1.25)),
      legend.title = element_text(size = rel(1.25)),
      legend.text = element_text(size = rel(1.25))
    )
```

This plot fits my eye much better. One piece of advice I got in undergraduate R class was to always apply something from ggthemes(). My personal favorite is theme_few(), which gives a stark white plot with a black outline. 

I like that the book advises people to default to using a colorblind-friendly color format. I personally perfer the "magmea" colorblind theme and with custom color ranges. 

I also feel that pretty much all default R settings for graphics are too small. When I think about communicating, I think of vizualizations as the stars of the show. For me, that means increasing the size of all text elements by 25% and setting sizes of `ggplot()` elements manually.  

For the `geom_line()`, I removed the SE band around the actual model. If I am not talking about a model I actually put work into, then it doesn't need to be there. I also made it black and extended it to the edges defined by `theme_few()`

Another piece of making vizualizations that are actually interesting is in better labeling. The title of the plot, if it has one, should highlight the most important finding. Here, the title points out that gentoos are distinctly bigger than the other two species. 


<!-- Visualizing Distributions 1.4 -->



The book moves into the second part of the first chapter with a discussion of **distribution**. Bar charts are used to show how redundant visual motifs can be used to keep coherence in a piece of writing. The bars do not need to be colored in the magma palette, but after the earlier scatterplots, one should try to keep visual themes consistent. Once Gentoo is defined as orange, it should stay that way. 

```{r}
#| label: fig-bar-species
#| fig-cap: ""
penguins |>
  ggplot(aes(x = species)) +
  geom_bar(aes(fill = species)) +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  labs(
    x = "Species",
    y = "Count",
    title = "Adelie is the most widespread species",
    caption = "Source: palmerpenguins"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme_few() +
  theme(
    axis.title = element_text(size = rel(1.25)),
    axis.text  = element_text(size = rel(1.25)),
    legend.position = "none" # species is already labeled on the x-axis
  )
```

Traditional statistics would tell you that a histogram like this is a useful way of showing distribution of a variable such as body mass.

```{r}
#| label: fig-histogram-mass
#| fig-cap: ""
penguins |>
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 200, fill = viridis::magma(1, begin = 0.5)) +
  labs(
    x = "Body Mass (g)",
    y = "Count",
    title = "Body mass has a slight bimodal shape",
    caption = "Source: palmerpenguins"
  ) +
  scale_x_continuous(expand = expansion(mult = 0.02)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme_few() +
  theme(
    axis.title = element_text(size = rel(1.25)),
    axis.text  = element_text(size = rel(1.25))
  )
```

however, I think in almost all cases, audiences would prefer to see a density plot. 

```{r}
#| label: fig-density-mass
#| fig-cap: ""
penguins |>
  ggplot(aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(aes(y = after_stat(count)), alpha = 0.3) +
  scale_color_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  labs(
    x = "Body Mass (g)",
    y = "Count",
    color = "Species",
    fill = "Species",
    title = "Gentoos are distinctly heavier than the other two species",
    caption = "Source: palmerpenguins"
  ) +
  scale_x_continuous(expand = expansion(mult = 0.02)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
  theme_few() +
  theme(
    axis.title  = element_text(size = rel(1.25)),
    axis.text   = element_text(size = rel(1.25)),
    legend.title = element_text(size = rel(1.25)),
    legend.text  = element_text(size = rel(1.25))
  )
```

#### Visualizing Relationships 1.5

##### Numerical and Categorical 1.5.1

A boxplot summarizes a numerical variable across categories. The box spans the middle 50% of the data (IQR), the line is the median, and the whiskers extend to 1.5× the IQR. Points beyond that are plotted individually as potential outliers.

```{r}
#| label: fig-boxplot-mass
#| fig-cap: ""
penguins |>
  ggplot(aes(x = species, y = body_mass_g)) +
  geom_boxplot(aes(fill = species)) +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  labs(
    x = "Species",
    y = "Body Mass (g)",
    title = "Gentoos are dramatically heavier than Adelie and Chinstrap",
    caption = "Source: palmerpenguins"
  ) +
  scale_y_continuous(labels = scales::label_number(),
                     expand = expansion(mult = 0.02)) +
  theme_few() +
  theme(
    axis.title = element_text(size = rel(1.25)),
    axis.text  = element_text(size = rel(1.25)),
    legend.position = "none" # species already on x-axis
  )
```

##### Two Categorical Variables 1.5.2

A stacked bar chart shows how two categorical variables relate. Using `position = "fill"` converts counts to proportions, making it easier to compare composition across groups regardless of group size.

```{r}
#| label: fig-bar-stacked
#| fig-cap: ""
penguins |>
  ggplot(aes(x = island, fill = species)) +
  geom_bar(position = "fill") +
  scale_fill_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  labs(
    x = "Island",
    y = "Proportion",
    fill = "Species",
    title = "Biscoe is dominated by Gentoos; Dream has no Gentoos at all",
    caption = "Source: palmerpenguins"
  ) +
  scale_y_continuous(expand = expansion(0)) +
  theme_few() +
  theme(
    axis.title  = element_text(size = rel(1.25)),
    axis.text   = element_text(size = rel(1.25)),
    legend.title = element_text(size = rel(1.25)),
    legend.text  = element_text(size = rel(1.25))
  )
```

##### Three or More Variables — Facets 1.5.4

When you want to show the same plot broken out by a categorical variable, `facet_wrap()` creates a panel for each group. This avoids overplotting and makes comparisons easy.

```{r}
#| label: fig-facet-island
#| fig-cap: ""
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), size = 3) +
  facet_wrap(~island) +
  scale_color_viridis_d(option = "magma", begin = 0.2, end = 0.8) +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species",
    title = "Biscoe penguins span the widest range of sizes",
    caption = "Source: palmerpenguins"
  ) +
  scale_x_continuous(expand = expansion(mult = 0.02)) +
  scale_y_continuous(labels = scales::label_number(),
                     expand = expansion(mult = 0.02)) +
  theme_few() 
```
