---
title: "Chapter 3: Data Transformation"
format: html
execute:
  warning: false
  message: false
  echo: false
---

Why chapter 3 and not chapter 2? Because Chapter 2 doesn't have code that is needed for future reference. If you've never written markdown code, it is useful. 

#### Primary lessons in Chapter 3
To answer the questions you have, you may need to change the form of your data. This may mean converting it to tidy format, creating new variables, and other things. 


```{r}
library(tidyverse)
library(nycflights13)
```

```{r}
glimpse(flights)
```

One of the packages containted in the **Tidyverse** newtowrk is *Dplyr* (a play on pliers), which contains the most-used tools for *transforming* data. 

```{r}
# dplyr functions do take standard logical operators
flights |> 
  filter(dep_delay > 120) 
```

```{r}
# %in% is used to check if an element is in a vector. In this case the vector is defined in the chunk itself as containing 1 or 2
flights |> 
  filter(month %in% c(1, 2))
```

```{r}
# distinct() finds unique rows. Left alone, it will just spit out unique rows. You can also define variables to find unique  combos
flights |> 
  distinct(origin, dest)
# this returns a two-column table of every combination of origin and destination that appears in the dataframe. 
```


